---
title: "AMR_deseq2"
author: "J. Meyer"
date: "2023-06-06"
output: html_document
---

Prepare data for DESeq2: 1. Count data should be formatted with genes in rows and samples in columns. 2. Metadata for the samples should be formatted with samples in rows and metadata in columns.


```{r, echo=FALSE}

counts <- read.table("20-samples-Bact-cds-counts.txt",sep="\t",header=TRUE)

# DeSeq is looking for only counts, without a first column of gene names, make the gene names row names.
mat <- counts[,-1]
rownames(mat) <- counts[,1]
# The Salmon counts have decimal places - round them up to prevent issues in deseq
counts2<-round(mat,digits=0)

# remove the disease samples from this analysis
counts2$AMR1D <- NULL
counts2$AMR2D <- NULL
counts2$AMR3D <- NULL
counts2$AMR5D <- NULL
counts2$AMR7D <- NULL
counts2$AMR8D <- NULL
counts2$AMR9D <- NULL

counts3 <-counts2[rowSums(counts2[])>0,] # remove empty rows
# use counts3 as input for DESeq2 
counts4 <- counts3
counts4$row_names <- row.names(counts4) # add back transcript.id as a column to merge with metadata
colnames(counts4)[14] <- 'transcript_id'
rgi <- read.table("rgi_out.txt",sep="\t",header=TRUE,quote="")
rgi_counts <-merge(rgi,counts4,by="transcript_id")
write.table(rgi_counts,"Untreated-vs-antibiotic_RGI-counts.txt",sep="\t",col.names=NA)

meta <-read.table("metadata.txt",sep="\t",header=TRUE)
# remove disease samples to compare just before and after antibiotic treatment
meta2 <-meta[meta$type != "disease",]
write.table(meta2,"metadata_13.txt",sep="\t",col.names=NA)

```

Summarize RGI genes by AMR Gene Family, Drug Class, and Resistance Mechanism in untreated vs antibiotic colonies; prepare data for DESeq2 analysis

```{r, echo=FALSE}
library(dplyr)
rgi_counts <- read.table("Untreated-vs-antibiotic_RGI-counts.txt",sep="\t",header=TRUE)
rgi_counts[1] <- NULL

#combine individual gene counts to get counts by AMR.Gene.Family
amrfam <- rgi_counts %>% group_by(AMR.Gene.Family) %>% summarize_at(vars(AMR1A:AMR10A), sum)
write.table(amrfam,"Untreated-vs-antibiotic_AMRgenefamily-counts.txt",sep="\t",col.names=NA)
#format for DESeq2
amrfam2 <- as.data.frame(amrfam)
rownames(amrfam2) <- amrfam2[,1]
amrfam2$AMR.Gene.Family <- NULL
# use amrfam2 as input for DESeq2

#combine individual gene counts to get counts by Drug.Class
drug <- rgi_counts %>% group_by(Drug.Class) %>% summarize_at(vars(AMR1A:AMR10A), sum)
write.table(drug,"Untreated-vs-antibiotic_DrugClass-counts.txt",sep="\t",col.names=NA)
#format for DESeq2
drug2 <- as.data.frame(drug)
rownames(drug2) <- drug2[,1]
drug2$Drug.Class <- NULL
# use drug2 as input for DESeq2

#combine individual gene counts to get counts by Resistance.Mechanism
mech <- rgi_counts %>% group_by(Resistance.Mechanism) %>% summarize_at(vars(AMR1A:AMR10A), sum)
write.table(mech,"Untreated-vs-antibiotic_ResistanceMechanism-counts.txt",sep="\t",col.names=NA)
#format for DESeq2
mech2 <- as.data.frame(mech)
rownames(mech2) <- mech2[,1]
mech2$Resistance.Mechanism <- NULL
# use mech2 as input for DESeq2

```

Run DESeq2 to test for differentially expressed *individual RGI genes* between coral colonies immediately before treatment ("untreated") and one day after amoxicillin treatment ("antibiotic").


```{r, echo=FALSE}
library(DESeq2)

meta <-read.table("metadata_13.txt",sep="\t",header=TRUE)
meta$X <- NULL
meta$type <-factor(meta$type)

# DESeq2 is looking for only counts, without a first column of gene names, make the gene names row names.
# This is testing all individual RGI genes
#construct DESEQDataSet Object
dds <- DESeqDataSetFromMatrix(countData=counts3, 
                              colData=meta, 
                              design=~type)
#see what the object looks like
dds

#now run DESeq function
dds <- DESeq(dds)
resultsNames(dds) # lists the coefficients
res <- results(dds, name="type_untreated_vs_antibiotic")
summary(res)

sink("DESeq2_summary_transcriptid.txt")
print(summary(res))
sink()

sink("DESeq2_results_transcriptid.txt")
print(res)
sink()

```


Run DESeq2 to test for differentially expressed *AMR gene families* between coral colonies immediately before treatment ("untreated") and one day after amoxicillin treatment ("antibiotic").


```{r, echo=FALSE}
library(DESeq2)

meta <-read.table("metadata_13.txt",sep="\t",header=TRUE)
meta$X <- NULL
meta$type <-factor(meta$type)

# DESeq2 is looking for only counts, without a first column of gene names, make the gene names row names.
# This is testing differential expression of AMR gene families
#construct DESEQDataSet Object
dds <- DESeqDataSetFromMatrix(countData=amrfam2, 
                              colData=meta, 
                              design=~type)
#see what the object looks like
dds

#now run DESeq function
dds <- DESeq(dds)
resultsNames(dds) # lists the coefficients
res <- results(dds, name="type_untreated_vs_antibiotic")
summary(res)

sink("DESeq2_results_AMRgenefamily.txt")
print(res)
sink()

sink("DESeq2_summary_AMRgenefamily.txt")
print(summary(res))
sink()
```

Run DESeq2 to test for differentially expressed *Drug Class* between coral colonies immediately before treatment ("untreated") and one day after amoxicillin treatment ("antibiotic").


```{r, echo=FALSE}
library(DESeq2)

meta <-read.table("metadata_13.txt",sep="\t",header=TRUE)
meta$X <- NULL
meta$type <-factor(meta$type)

# DESeq2 is looking for only counts, without a first column of gene names, make the gene names row names.
# This is testing differential expression of AMR gene families
#construct DESEQDataSet Object
dds <- DESeqDataSetFromMatrix(countData=drug2, 
                              colData=meta, 
                              design=~type)
#see what the object looks like
dds

#now run DESeq function
dds <- DESeq(dds)
resultsNames(dds) # lists the coefficients
res <- results(dds, name="type_untreated_vs_antibiotic")
summary(res)

sink("DESeq2_results_DrugClass.txt")
print(res)
sink()

sink("DESeq2_summary_DrugClass.txt")
print(summary(res))
sink()
```

Run DESeq2 to test for differentially expressed *Resistance Mechanism* between coral colonies immediately before treatment ("untreated") and one day after amoxicillin treatment ("antibiotic").


```{r, echo=FALSE}
library(DESeq2)

meta <-read.table("metadata_13.txt",sep="\t",header=TRUE)
meta$X <- NULL
meta$type <-factor(meta$type)

# DESeq2 is looking for only counts, without a first column of gene names, make the gene names row names.
# This is testing differential expression of AMR gene families
#construct DESEQDataSet Object
dds <- DESeqDataSetFromMatrix(countData=mech2, 
                              colData=meta, 
                              design=~type)
#see what the object looks like
dds

#now run DESeq function
dds <- DESeq(dds)
resultsNames(dds) # lists the coefficients
res <- results(dds, name="type_untreated_vs_antibiotic")
summary(res)

sink("DESeq2_results_ResistanceMechanism.txt")
print(res)
sink()

sink("DESeq2_summary_ResistanceMechanism.txt")
print(summary(res))
sink()
```


Since nothing is differentially expressed between untreated and antibiotic treated colonies, plot summary of the types of AMR genes that were expressed. I will use Resistance Mechanism since there are 11 types in this category. (Drug Class = 140 types, AMR gene families = 371 types)

```{r, echo=FALSE}
library(ggplot2)
library(reshape2)

meta <-read.table("metadata_13.txt",sep="\t",header=TRUE)
meta$X <- NULL
mech <-read.table("Untreated-vs-antibiotic_ResistanceMechanism-counts.txt",sep="\t",header=TRUE)
mech$X <- NULL
#calculate relative abundance from counts
mech_RA <- mech[,-1]/colSums(mech[,-1]) * 100
mech_RA$Resistance.Mechanism <- mech$Resistance.Mechanism

#merge metadata and count dataframes - start by converting mech from wide to long format
mech_long <- melt(mech_RA, id.vars=c("Resistance.Mechanism"))
colnames(mech_long)[colnames(mech_long) == "variable"] <- "sample"
colnames(mech_long)[colnames(mech_long) == "value"] <- "proportion"
resist <-merge(meta, mech_long, "sample")
write.table(resist,"Untreated-vs-antibiotic_ResistanceMechanism-RA_combine_other.txt",sep="\t",col.names=NA)


#in excel, I changed categories with multiple mechanisms listed into "other" to reduce total number of displayed categories (and get rid of super long names in legend that prevented successful plotting)
other <-read.table("Untreated-vs-antibiotic_ResistanceMechanism-RA_combine_other.txt",sep="\t",header=TRUE)
other$Resistance.Mechanism <-factor(other$Resistance.Mechanism, levels = c("antibiotic efflux","antibiotic inactivation","antibiotic target alteration","antibiotic target protection","antibiotic target replacement","reduced permeability to antibiotic","other"))
mycolors <-c("antibiotic efflux"="#E69F00","antibiotic inactivation"="#D55E00","antibiotic target alteration"="#56B4E9","antibiotic target protection"="#0072B2","antibiotic target replacement"="#009E73","reduced permeability to antibiotic"="#CC79A7","other"="#999999")

#plot as stacked bars
pdf("resistance_mechanisms.pdf", height=4)
p1<-ggplot(other, aes(fill=Resistance.Mechanism, x=sample, y=proportion))+
  geom_bar(position="fill",stat="identity")+
  facet_grid(. ~ type, scales="free", space="free")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90))+
  scale_fill_manual(values=mycolors)
p1
dev.off()

#group colonies to look at before/after
pdf("resistance_mechanisms_v2.pdf", height=4)
p1<-ggplot(other, aes(fill=Resistance.Mechanism, x=sample, y=proportion))+
  geom_bar(position="fill",stat="identity")+
  facet_grid(. ~ colony, scales="free", space="free")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90))+
  scale_fill_manual(values=mycolors)
p1
dev.off()

```

Plot the resistance mechanism "antibiotic inactivation" which includes beta-lactamases


```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
rgi_counts <- read.table("Untreated-vs-antibiotic_RGI-counts.txt",sep="\t",header=TRUE)
rgi_counts[1] <- NULL

#subset = just "antibiotic inactivation" which includes beta-lactamases
inact <- rgi_counts[which(rgi_counts$Resistance.Mechanism=='antibiotic inactivation'),]

inact2 <- inact %>% group_by(AMR.Gene.Family) %>% summarize_at(vars(AMR1A:AMR10A), sum)
write.table(inact2,"Untreated-vs-antibiotic_AntibioticInactivation-counts.txt",sep="\t",col.names=NA)
#calculate relative abundance from counts
inact2_RA <- inact2[,-1]/colSums(inact2[,-1]) * 100
inact2_RA$AMR.Gene.Family <- inact2$AMR.Gene.Family
inact3 = inact2_RA[, c("AMR.Gene.Family", names(inact2_RA)[names(inact2_RA) != "AMR.Gene.Family"])]
write.table(inact3,"Untreated-vs-antibiotic_AntibioticInactivation-RA.txt",sep="\t",col.names=NA)

#in excel remove all rows that are not beta-lactamases 215 out of 260 inactivation gene families are beta-lactamases
beta <- read.table("Untreated-vs-antibiotic_AntibioticInactivation-RA_v2.txt",sep="\t",header=TRUE)

#create a sum column to find most abundant beta-lactamase gene families then sort by abundance before plotting
beta$total <- rowSums(beta[,2:14])
beta <- beta[order(beta$total, decreasing = TRUE),]
top25 <-head(beta, 25)
write.table(top25,"Untreated-vs-antibiotic_AntibioticInactivation-RA_top25.txt",sep="\t",col.names=NA)

top10 <-head(inact4, 10)
write.table(top10,"Untreated-vs-antibiotic_AntibioticInactivation-RA_top10.txt",sep="\t",col.names=NA)


#merge metadata and count dataframes - start by converting mech from wide to long format
top25_long <- melt(top25, id.vars=c("AMR.Gene.Family"))
colnames(top25_long)[colnames(top25_long) == "variable"] <- "sample"
colnames(top25_long)[colnames(top25_long) == "value"] <- "proportion"
meta <-read.table("metadata_13.txt",sep="\t",header=TRUE)
meta$X <- NULL
inact5 <-merge(meta, top25_long, "sample")

#plot as stacked bars
#pdf("inactivation.pdf", height= 11, width=8.5)
ggplot(inact5, aes(fill=AMR.Gene.Family, x=sample, y=proportion))+
  geom_bar(position="fill",stat="identity")+
  facet_grid(. ~ type, scales="free", space="free")+
  theme(legend.position="bottom")
#dev.off()


#merge metadata and count dataframes - start by converting mech from wide to long format
top10_long <- melt(top10, id.vars=c("AMR.Gene.Family"))
colnames(top10_long)[colnames(top10_long) == "variable"] <- "sample"
colnames(top10_long)[colnames(top10_long) == "value"] <- "proportion"
meta <-read.table("metadata_13.txt",sep="\t",header=TRUE)
meta$X <- NULL
inact6 <-merge(meta, top10_long, "sample")


#plot as stacked bars
inact6$AMR.Gene.Family <-factor(inact6$AMR.Gene.Family, levels = c("OXA beta-lactamase","nitroimidazole reductase","tetracycline inactivation enzyme","chloramphenicol acetyltransferase (CAT)","streptogramin vgb lyase","ANT(3'')","macrolide phosphotransferase (MPH)","ADC beta-lactamases pending classification for carbapenemase activity","CfiA beta-lactamase","MOX beta-lactamase"))

mycolors2 <-c("antibiotic efflux"="#E69F00","antibiotic inactivation"="#D55E00","antibiotic target alteration"="#56B4E9","antibiotic target protection"="#0072B2","antibiotic target replacement"="#009E73","reduced permeability to antibiotic"="#CC79A7","other"="#999999")


pdf("inactivation_top10.pdf")
ggplot(inact6, aes(fill=AMR.Gene.Family, x=sample, y=proportion)) +geom_bar(position="fill",stat="identity")+facet_grid(. ~ type, scales="free", space="free")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90)) 
#+ theme(legend.position="bottom")
dev.off()

# to do: black and white theme, better fill colors, 90 degree shift of x-axis labels
# reduce hybrid categories to "other"



```

